---
title: "Introduction to cori.data.pep"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to cori.data.pep}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval    = TRUE,
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 6,
  out.width  = "100%"
)
```

`cori.data.pep` provides county-level population estimates from the U.S. Census Bureau's Population Estimates Program (PEP), processed and stored in CORI's S3 data lake. Data cover 2000–present at annual frequency.

## Variables

```{r codebook}
library(cori.data.pep)

get_pep_codebook() |>
  dplyr::select(variable, label, coverage = notes) |>
  knitr::kable()
```

## Reading data

All data are returned in long format: one row per `geoid / year / variable`.

```{r read-all}
# Total population, all counties, all years
pop <- read_pep_from_s3(variables = "population")
dplyr::glimpse(pop)
```

Filter to specific variables, years, or counties:

```{r read-filtered}
# Migration components for West Virginia counties
wv_mig <- read_pep_from_s3(
  variables = c("domestic_mig", "international_mig", "net_mig"),
  geoids    = grep("^54", unique(pop$geoid), value = TRUE),
  years     = 2010:2025
)

dplyr::glimpse(wv_mig)
```

## Population trend — Grafton County, NH

Grafton County, NH (GEOID `33009`) illustrates how to build a simple time-series chart using `cori.charts`.

```{r grafton-pop, fig.alt="Line chart of Grafton County NH total population 2000-2025"}
library(ggplot2)
library(cori.charts)

load_fonts()

grafton_pop <- read_pep_from_s3(
  variables = "population",
  geoids    = "33009",
  years     = 2000:2025
)

ggplot(grafton_pop, aes(x = year, y = value)) +
  geom_line(color = cori_colors[["Dark Green"]], linewidth = 1.2) +
  geom_point(color = cori_colors[["Dark Green"]], size = 2) +
  scale_x_continuous(breaks = seq(2000, 2025, by = 5)) +
  scale_y_continuous(labels = scales::label_comma()) +
  theme_cori() +
  labs(
    title    = "Grafton County, NH population, 2000\u20132025",
    subtitle = "Total resident population estimate",
    x        = NULL,
    y        = NULL,
    caption  = "Source: CORI analysis of U.S. Census Bureau Population Estimates Program (PEP)."
  )
```
